<<<<<<< HEAD
---
title: "Final Project"
author: "Ikjoo Oh"
date: "`r Sys.Date()`"
output: pdf_document
---

=======
---
title: "Final Project"
author: "Ikjoo Oh"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
wages_by_education <- read.csv("wages_by_education.csv")

# Load packages
suppressPackageStartupMessages(library(tidyverse))
```



```{r}
# Data of White people
wbe_white <- wages_by_education %>%
  select(year,white_less_than_hs:white_advanced_degree)
```

```{r}
# Data of Black people
wbe_black <- wages_by_education %>%
  select(year,black_less_than_hs:black_advanced_degree)
```

```{r}
# Data of Hispanic People
wbe_hispanic <- wages_by_education %>%
  select(year,hispanic_less_than_hs:hispanic_advanced_degree)
```

```{r}
# Black people divided in their level of education
wbe_black_education <-wbe_black %>%
pivot_longer(cols =2:black_advanced_degree, names_to = 'educational_level',
             values_to = 'wages')
```

```{r}
wbe_black_education %>%
  ggplot()+
  geom_line(mapping = aes(x = year, 
                          y = wages,
                          color = educational_level)

)
```




```{r}
# White people divided in their level of education
wbe_white_education <-wbe_white %>%
pivot_longer(cols =2:white_advanced_degree, names_to = 'educational_level',
             values_to = 'wages')
```



```{r}
wbe_white_education %>%
  ggplot()+
  geom_line(mapping = aes(x = year, 
                          y = wages,
                          color = educational_level)

)
```



```{r}
# Hispanic people divided in their level of education
wbe_hispanic_education <-wbe_hispanic %>%
pivot_longer(cols =2:hispanic_advanced_degree, names_to
             ='educational_level',
             values_to = 'wages')
```

```{r}
wbe_hispanic_education %>%
  ggplot()+
  geom_line(mapping = aes(x = year, 
                          y = wages,
                          color = educational_level)

)
```
